Womb for One More? Fertility Expansion and the Limits of Early Health Investment: Empirical Evidence on the Quantity-Quality Trade-off from a Contraceptive Ban in Urban Philippines

### 2025-07-27 docs/paper.Rmd
revise theory and empirical sections in paper draft
Expanded and clarified the literature review, conceptual framework, and empirical strategy in the paper. Standardized terminology from 'quantity–quality' to 'quantity-quality', added new references and recent evidence, and restructured sections for improved flow. Updated title and subtitle for consistency across Rmd, tex, and PDF outputs.

---
title: "Womb for One More? Fertility Expansion and the Limits of Early Health Investment"
subtitile: "Empirical Evidence on the Quantity-Quality Trade-off from a Contraceptive Ban in Urban Philippines"
short: "The Q-Q Trade-off in the Philippine Context"
journal: "" # AER, AEJ, PP, JEL
month: "`r format(Sys.Date(), '%m')`"
year: "`r format(Sys.Date(), '%Y')`"
jel:
  - J13  # Fertility; Childbearing; Family Planning; Population Policy
  - I15  # Health and Economic Development
  - O15  # Human Resources; Human Development; Income Distribution; Migration
keywords:
  - fertility shocks
  - child nutrition
  - quantity-quality trade-off
  - urban poverty
  - Philippines
author:
  - name: Erika Salvador
    firstname: Erika
    surname: Salvador
    email: esalvador28@amherst.edu
    affiliation: Amherst College
  - name: Caroline Theoharides
    firstname: Caroline
    surname: Theoharides
    email: ctheoharides@amherst.edu
    affiliation: Amherst College
acknowledgements: |
  This research was made possible by the Schupf Fellowship at Amherst College. I am indebted to my faculty adviser, Professor Caroline Theoharides, for her mentorship and support throughout the development of this project. I also wish to thank Faculty Director Professor Amelie Hastie and the campus partners whose institutional support enabled this work. I am further grateful to the Departments of Economics and Mathematics & Statistics for their academic support. The views expressed and any errors contained herein are entirely my own responsibility.
abstract: |
  This paper examines whether increased fertility, driven by reduced access to modern contraception, affects early-life health outcomes in low-income urban households. I exploit a natural experiment created by a 2000 policy in Manila, Philippines—Executive Order No. 003—which effectively banned the provision of modern contraceptives in all city-run health facilities. Using a difference-in-differences design and nationally representative data from the Philippine Demographic and Health Surveys, I estimate the policy’s impact on child health by comparing outcomes in Manila before and after the ban to those in comparable urban areas unaffected by the policy. The outcomes of interest include birth weight (in grams), the mother’s perceived size of the child at birth, and receipt of key childhood vaccinations.
output: 
  rticles::aea_article:
    keep_tex: true
    latex_engine: pdflatex
bibliography: references.bib
---

@@

The quantity–quality (Q–Q) theory, a central idea in modern family economics,  holds that parents face a trade-off between the number of children and the "quality" of investment—such as education or health—they can provide to each. Quality in this context refers to the human capital of each child: attributes like education, health, and nutrition that enhance a child’s future productivity and well-being. The genesis of this idea traces back to Gary Becker’s seminal work around 1960, which for the first time treated children as economic goods subject to parental choice and budget constraints \citep{becker1960economic}. Becker argued that as families become wealthier, they may not simply want more children, but rather better-raised children, much as a household might prefer a higher-quality car or house over a greater quantity of them. This proposition led to a formal theory in which increases in income or changes in economic conditions cause parents to substitute child quality for quantity, consistent with historical patterns of lower fertility and higher educational attainment during economic development \citep{galor2000population}.

@@

Becker and Lewis (1973) formalize the mutually reinforcing nature of the quantity–quality cost structure. An increase in $n$ raises the total cost required to sustain a given level of $q$ for each child, while a higher level of $q$ raises the marginal cost associated with having an additional child. For example, allocating more resources to education or health per child increases the financial burden of expanding family size. This interdependence links the two decisions directly. The household cannot choose $n$ and $q$ in isolation; each choice alters the marginal cost of the other.

@@

In many dynastic models, child quality appears indirectly, often through the child’s future human capital or income. One variant assumes parents value the aggregate human‐capital stock of their children. This specification, combined with altruism, produces a similar trade‐off: concentrating resources in fewer children raises each child’s human capital, which raises the dynasty’s long‐run welfare. These intergenerational extensions link micro‐level fertility decisions to macroeconomic outcomes. By the late 1980s, work by Becker, Barro, and others had recast fertility as an endogenous choice that interacts with capital accumulation, income distribution, and policy. This laid the foundation for unified growth theories, which view the quantity–quality mechanism as central to demographic transition and long-run development.

@@

The unified growth literature places the Q–Q model within a more general account of economic and demographic change. In this framework, higher income or stronger returns to child quality reduce fertility and help shift economies from stagnation toward sustained growth. Several mechanisms support this transition, such as a fall in child labor, a drop in child mortality, and a shift in parental priorities. These models explain not only the presence of a quantity–quality trade-off but also its rising influence at a specific point in history. The evidence supports these claims: countries that saw earlier increases in returns to education experienced earlier fertility decline, while delays in reforms, such as public education or health access, corresponded to prolonged high fertility. As a result, the Q–Q mechanism forms a key component of unified growth theory.

@@

Contemporary research has further refined the quantity–quality model by relaxing some of its initial simplifying assumptions. Three important extensions involve (1) capital market imperfections that constrain parents’ ability to invest in child quality, (2) intra-household conflict and bargaining between mothers and fathers over fertility choices, and (3) recognition that child quality is multi-dimensional, which extends beyond schooling to include health, nutrition, and other facets of human capital.

To begin, I examine how credit constraints can give rise to poverty traps. The canonical quantity–quality (Q–Q) framework assumes that parents can reallocate resources freely; borrowing against future earnings to finance schooling or health investments whenever the expected return is high. In practice, especially in low-income settings, credit markets function imperfectly: poor households typically cannot secure loans to cover children’s education or medical costs even when such investments would yield substantial future gains. This market failure magnifies the Q–Q trade-off. Becker, Lewis, and Willis–already noted by \citet{grawe2008quality} as emphasizing “resource limitations”–implicitly recognized this issue, but contemporary models make it explicit by imposing a borrowing constraint. Parents must fund childrearing from current income alone; they cannot collateralize a child’s future wages to pay today’s school fees. Consequently, when income is low, each additional birth directly reduces the attainable quality per child, potentially trapping families in a low-income–high-fertility equilibrium.

Formal models show that “the quality–quantity trade-off arises from a binding credit constraint that prevents parents from borrowing against future child income.” Empirical work supports this mechanism. \citet{kremer2002income} and \citet{de2003inequality} document that countries facing tighter liquidity constraints tend to display higher fertility and lower educational attainment, consistent with liquidity-constrained parents favoring quantity over quality. Cross-country evidence likewise indicates that where financial frictions are more severe, the negative correlation between fertility and schooling is stronger. Theoretically, introducing a borrowing limit can generate multiple steady states: one with low fertility and high investment when incomes suffice to cover quality costs, and another with high fertility and minimal investment when they do not. Policy instruments such as education subsidies or conditional cash transfers effectively relax these constraints, which nudge households toward the low-fertility, high-investment equilibrium. In short, incorporating credit market imperfections deepens the explanatory power of the Q-Q model: economic growth alone may not reduce fertility if households remain too cash-poor to afford schooling, whereas targeted quality-enhancing transfers can catalyze both demographic and human-capital transitions.

A further refinement of the quantity–quality (Q–Q) model considers the question of *who* within the household makes fertility and child investment decisions. The original Beckerian framework adopts a unitary model of the family, and assumes a single utility function and complete agreement between spouses over optimal fertility \( n \) and child quality \( q \). In practice, however, empirical evidence reveals significant heterogeneity in preferences between household members along gender lines. According to \citet{oppenheim1987impact, thomas1990intra}, for instance, men often desire more children than women and may differ in their willingness to invest in each child’s education or health. These discrepancies have motivated game-theoretic models of intra-household bargaining, in which fertility and investment outcomes reflect the relative influence of each parent’s preferences.

@@

The original Q–Q models often used a single catch-all variable for child quality, typically thought of as education or “expenditure per child.” Recent work emphasizes that child quality is multi-faceted, and that parents make trade-offs along several dimensions of investment—cognitive development, health, nutrition, etc. This is particularly salient in developing countries, where basic health and nutrition are pressing concerns alongside schooling.

@@

Beyond survival, nutritional status is a critical indicator of quality in many low-income settings. Parents often face choices about how to allocate food or medical care, especially under resource scarcity. Empirical evidence supports the Q–Q model in this context: children with many siblings often exhibit poorer nutritional outcomes. A recent study using Vietnam Young Lives data finds that an additional sibling reduces height-for-age and weight-for-age z-scores by about 0.3 standard deviations on average \citep{chen2021population}  This effect is substantial, suggesting that children in larger families tend to be shorter and lighter than their peers from smaller families, likely due to the more limited distribution of healthcare and parental attention

Similar patterns emerge in sub-Saharan Africa \citep{bishwakarma2019first} and South Asia , where first-born children experience slower growth once younger siblings arrive—again pointing to intrahousehold trade-offs. Theoretically, one can extend the Becker model with a nutrition production function, where a child’s health \( H \) depends on food \( F \) and medical care \( M \), such that:

@@

Early empirical work in the Philippines provided suggestive support for this theory. For example, an influential study by \citet{horton1986child} used Philippine household data to jointly examine fertility and child nutrition and treated nutritional status as a measure of child quality. Horton found that better-educated mothers and fathers tended to choose smaller families and achieved better-nourished children–evidence of substitution away from “quantity” toward child “quality.” Notably, she observed strong birth-order effects on nutrition (first-born children faring better than later-born), which hint that parents may not evenly distribute quality investments among all children. This early work indicated a quantity–quality (Q–Q) trade-off in Philippine families, though it largely documented correlations rather than definitive causation.

@@

The IV estimates confirmed a significant negative causal effect of higher fertility on educational outcomes. \citet{orbeta2010number} found that each additional child in the household reduced the proportion of school-age children (6–24 years) attending school by roughly 19% of the baseline attendance rate. The trade-off was especially pronounced at higher education levels: for example, the estimated drop in school attendance was about 26% at the secondary level and 57% at the tertiary level for each additional sibling. These are sizable effects, implying that children from large families are substantially less likely to remain in school, presumably due to tighter household budget constraints or diluted parental attention. Moreover, the burden of the trade-off appeared regressive: Orbeta’s results showed much larger schooling deficits from an extra child in poorer households than in richer ones. For instance, in the poorest quintile, an additional sibling reduced school attendance by an estimated 24% (for ages 6–24), compared to a 16% reduction in the richest quintile. This regressive pattern aligns with Becker’s theory that resource constraints bind more tightly for low-income families, which makes the Q–Q trade-off more acute. In summary, Orbeta’s study – the first in the Philippines to account for fertility endogeneity – provides robust evidence that increases in family size cause significant declines in child educational attainment and validates the quantity–quality trade-off in this context. Large family size thus emerges as one mechanism contributing to poverty, by impeding children’s human capital accumulation in the Philippines.

@@

The Philippine experience is echoed in other ASEAN countries, where researchers have probed the quantity–quality trade-off with diverse outcomes and methods. In Vietnam, for example, rapid fertility decline alongside rising education led to questions about a Q–Q mechanism. \citet{anh1998family} documented a negative correlation between family size and children’s school enrollment in Vietnam, though their analysis could not fully establish causality.

@@

Nonetheless, the prevailing evidence from the Philippines and its regional neighbors strongly supports Becker’s quantity–quality conjecture. As families have fewer children, they appear to invest more in each child’s education—investments crucial for human capital development and economic growth. Conversely, high-fertility households risk under-investing per child, which reinforces cycles of poverty and inequality. This literature thus provides an important empirical foundation for policies in the Philippines and ASEAN—from family planning programs to education subsidies—that aim to ease the quantity–quality trade-off and help families achieve both manageable size and better outcomes for the next generation.

@@

This study builds upon the foundational framework of the quantity–quality (Q–Q) trade-off, as articulated in \citet{becker1960economic} seminal work and later refined by \citet{becker1973interaction}. Within this theoretical model, fertility and child quality are jointly determined under a full-income constraint, such that any increase in the number of children—absent a commensurate expansion in resources—necessitates a reduction in the average investment per child. While prior empirical applications of this framework have largely centered on educational attainment, child quality may also be understood along biological dimensions, especially along nutritional status, which emerges earlier in the life course and may be less subject to later remediation.

The current analysis focuses exclusively on early-life nutrition as a proxy for child quality. Nutrition, unlike cognitive or educational outcomes, is characterized by low substitutability and a narrow temporal window during which investments are most productive. The trade-off is thus expected to manifest strongly in settings where fertility rises exogenously while household resources remain constrained. In such environments, families facing an unexpected increase in sibship size may be forced to reallocate scarce food, time, and health inputs across a larger number of dependents, potentially resulting in lower nutritional attainment for each child.

@@

In the presence of a fertility shock, holding \( I \) constant, the average \(\kappa_j\) must fall unless offset by changes in parental behavior or external transfers. If nutritional investments per child decline, and if such inputs are critical for growth, observable deficits in height-for-age (HAZ) or weight-for-height (WHZ) may emerge.
The first hypothesis emerging from this framework is that an exogenous increase in

fertility leads to a reduction in per-child nutritional investment, particularly in low-income, urban households where slack in the family budget is limited. A second, related hypothesis posits that this reduction in inputs translates into measurable nutritional deficits among young children, specifically in the form of lower anthropometric outcomes such as height-for-age (HAZ) and weight-for-height (WHZ) z-scores. These two hypotheses—concerning, respectively, the behavioral reallocation of resources and its biological consequence—jointly seek to identify a causal link between fertility shocks and early-life deprivation.

@@

This study constructs a harmonized, child-level analytic dataset using five rounds of the Philippine Demographic and Health Survey (DHS): 1993, 1998, 2003, 2008, and 2013. The DHS is a nationally representative, cross-sectional survey series conducted by the Philippine Statistics Authority in collaboration with ICF International. Each round follows a stratified two-stage cluster design and collects standardized information on child health, fertility, maternal characteristics, and household infrastructure. For each wave, I merge the child-level (KR), mother-level (IR), and household-level (HR) files using cluster, household, and maternal identifiers, ensuring that each observation corresponds to a live birth occurring within five years of the survey. The analytic sample is restricted to children aged 0–59 months to standardize developmental exposure windows and conform to the World Health Organization (WHO) reference growth standards.

The primary outcomes are height-for-age (HAZ) and weight-for-height (WHZ) z-scores, internationally recognized proxies for chronic and acute malnutrition, respectively. These anthropometric indicators are cleaned following DHS protocols: biologically implausible values are excluded, and remaining z-scores are rescaled from hundredths to continuous units. In addition to anthropometry, the child record includes a suite of health-related variables. These include indicators for whether the child has received basic vaccinations—such as BCG, DPT, polio, and measles—as well as whether the child experienced episodes of diarrhea, fever, or respiratory illness in the two weeks prior to the survey. These supplementary health measures serve as intermediate outcomes in secondary analyses to capture short-run morbidity and preventive health investment.

Maternal-level information is drawn from the individual recode file. Core covariates include maternal age, completed years of education, and total number of children ever born (parity), each of which captures socioeconomic status and fertility history. I additionally include maternal body mass index (BMI) and anemia status, which serve as both controls and potential mediators of child health. These indicators reflect the mother’s nutritional status and physiological capacity to support healthy fetal and postnatal development. In robustness checks, I examine whether declines in maternal nutritional status following the fertility shock account for observed changes in child anthropometry.

Household-level characteristics are used to control for environmental and infrastructural determinants of health. Specifically, I include household size, urban residence, wealth quintile, access to electricity, type of toilet facility, whether the toilet is shared, and source of drinking water. These variables proxy for household-level resources and exposure to disease environments. In extension analyses, I interact treatment status with these environmental indicators to assess whether the nutritional consequences of the fertility shock are magnified under infrastructural deprivation.

The treatment is defined by exposure to Executive Order No. 003, which was issued by the Manila City government in January 2000 and effectively suspended the public distribution of modern contraceptives. This policy change resulted in a localized increase in fertility that was not mirrored in surrounding municipalities within the National Capital Region (NCR). Children born in Manila City after the implementation of the policy are classified as treated, while children in other NCR cities serve as the control group. This spatially and temporally bounded intervention provides quasi-experimental variation that facilitates identification of causal effects.

The empirical approach adopts a difference-in-differences framework that compares trends in child nutritional outcomes between treated and control cities before and after the policy change. The identifying assumption is that, absent the policy, nutritional trajectories in Manila and non-Manila NCR cities would have evolved in parallel. This assumption is supported by visual inspection and pre-trend placebo tests using the 1993 and 1998 waves. The 1998 DHS serves as the primary pre-policy baseline, while the 2003 survey captures short-run effects. The 2008 and 2013 waves are used in robustness checks to assess medium- and long-run persistence of nutritional deficits. All specifications include city and birth-year fixed effects, and standard errors are clustered at the city-year level to account for within-cluster correlation.

The empirical strategy exploits the 2000 issuance of Executive Order No. 003, which effectively banned the distribution of modern contraceptives in public health centers across Manila City. This abrupt and localized withdrawal of contraceptive access resulted in a fertility increase concentrated within the city, but did not affect surrounding municipalities within the National Capital Region (NCR). Children born in Manila following the policy are considered exposed to a fertility shock, while children in other NCR cities over the same period serve as the control group. The analytical sample is restricted to children under five years old at the time of each survey round to ensure accurate measurement of early-life nutritional outcomes.

I estimate the following difference-in-differences specification:

$$ \text{Nutrition}_{ict} = \alpha + \beta \cdot (\text{Post}_t \times \text{Manila}_c) + X_{ict}'\gamma + \lambda_c + \tau_t + \varepsilon_{ict} $$

>>>>

---
title: "Womb for One More? Fertility Expansion and the Limits of Early Health Investment: Empirical Evidence on the Quantity-Quality Trade-off from a Contraceptive Ban in Urban Philippines"
short: "The Q-Q Trade-off in the Philippine Context"
journal: "" # AER, AEJ, PP, JEL
month: "`r format(Sys.Date(), '%m')`"
year: "`r format(Sys.Date(), '%Y')`"
jel:
  - J13  # Fertility; Childbearing; Family Planning; Population Policy
  - I15  # Health and Economic Development
  - O15  # Human Resources; Human Development; Income Distribution; Migration
keywords:
  - fertility shocks
  - child nutrition
  - quantity-quality trade-off
  - urban poverty
  - Philippines
author:
  - name: Erika Salvador
    firstname: Erika
    surname: Salvador
    email: esalvador28@amherst.edu
    affiliation: Amherst College
  - name: Caroline Theoharides
    firstname: Caroline
    surname: Theoharides
    email: ctheoharides@amherst.edu
    affiliation: Amherst College
acknowledgements: |
  This research was made possible by the Schupf Fellowship at Amherst College. I am indebted to my faculty adviser, Professor Caroline Theoharides, for her mentorship and support throughout the development of this project. I also wish to thank Faculty Director Professor Amelie Hastie and the campus partners whose institutional support enabled this work. I am further grateful to the Departments of Economics and Mathematics & Statistics for their academic support. The views expressed and any errors contained herein are entirely my own responsibility.
abstract: |
  This paper examines whether increased fertility, driven by reduced access to modern contraception, affects early-life health outcomes in low-income urban households. I exploit a natural experiment created by a 2000 policy in Manila, Philippines—Executive Order No. 003—which effectively banned the provision of modern contraceptives in all city-run health facilities. Using a difference-in-differences design and nationally representative data from the Philippine Demographic and Health Surveys, I estimate the policy’s impact on child health by comparing outcomes in Manila before and after the ban to those in comparable urban areas unaffected by the policy. The outcomes of interest include birth weight (in grams), the mother’s perceived size of the child at birth, and receipt of key childhood vaccinations.
output: 
  rticles::aea_article:
    keep_tex: true
    latex_engine: pdflatex
bibliography: references.bib
---

@@

The quantity-quality (Q–Q) theory, a central idea in modern family economics,  holds that parents face a trade-off between the number of children and the "quality" of investment—such as education or health—they can provide to each. Quality in this context refers to the human capital of each child: attributes like education, health, and nutrition that enhance a child’s future productivity and well-being. The genesis of this idea traces back to Gary Becker’s seminal work around 1960, which for the first time treated children as economic goods subject to parental choice and budget constraints \citep{becker1960economic}. Becker argued that as families become wealthier, they may not simply want more children, but rather better-raised children, much as a household might prefer a higher-quality car or house over a greater quantity of them. This proposition led to a formal theory in which increases in income or changes in economic conditions cause parents to substitute child quality for quantity, consistent with historical patterns of lower fertility and higher educational attainment during economic development \citep{galor2000population}.

@@

Becker and Lewis (1973) formalize the mutually reinforcing nature of the quantity-quality cost structure. An increase in $n$ raises the total cost required to sustain a given level of $q$ for each child, while a higher level of $q$ raises the marginal cost associated with having an additional child. For example, allocating more resources to education or health per child increases the financial burden of expanding family size. This interdependence links the two decisions directly. The household cannot choose $n$ and $q$ in isolation; each choice alters the marginal cost of the other.

@@

In many dynastic models, child quality appears indirectly, often through the child’s future human capital or income. One variant assumes parents value the aggregate human‐capital stock of their children. This specification, combined with altruism, produces a similar trade‐off: concentrating resources in fewer children raises each child’s human capital, which raises the dynasty’s long‐run welfare. These intergenerational extensions link micro‐level fertility decisions to macroeconomic outcomes. By the late 1980s, work by Becker, Barro, and others had recast fertility as an endogenous choice that interacts with capital accumulation, income distribution, and policy. This laid the foundation for unified growth theories, which view the quantity-quality mechanism as central to demographic transition and long-run development.

@@

The unified growth literature places the Q–Q model within a more general account of economic and demographic change. In this framework, higher income or stronger returns to child quality reduce fertility and help shift economies from stagnation toward sustained growth. Several mechanisms support this transition, such as a fall in child labor, a drop in child mortality, and a shift in parental priorities. These models explain not only the presence of a quantity-quality trade-off but also its rising influence at a specific point in history. The evidence supports these claims: countries that saw earlier increases in returns to education experienced earlier fertility decline, while delays in reforms, such as public education or health access, corresponded to prolonged high fertility. As a result, the Q–Q mechanism forms a key component of unified growth theory.

@@

Contemporary research has further refined the quantity-quality model by relaxing some of its initial simplifying assumptions. Three important extensions involve (1) capital market imperfections that constrain parents’ ability to invest in child quality, (2) intra-household conflict and bargaining between mothers and fathers over fertility choices, and (3) recognition that child quality is multi-dimensional, which extends beyond schooling to include health, nutrition, and other facets of human capital.

To begin, I examine how credit constraints can give rise to poverty traps. The canonical quantity-quality (Q–Q) framework assumes that parents can reallocate resources freely; borrowing against future earnings to finance schooling or health investments whenever the expected return is high. In practice, especially in low-income settings, credit markets function imperfectly: poor households typically cannot secure loans to cover children’s education or medical costs even when such investments would yield substantial future gains. This market failure magnifies the Q–Q trade-off. Becker, Lewis, and Willis–already noted by \citet{grawe2008quality} as emphasizing “resource limitations”–implicitly recognized this issue, but contemporary models make it explicit by imposing a borrowing constraint. Parents must fund childrearing from current income alone; they cannot collateralize a child’s future wages to pay today’s school fees. Consequently, when income is low, each additional birth directly reduces the attainable quality per child, potentially trapping families in a low-income–high-fertility equilibrium.

Formal models show that “the quality-quantity trade-off arises from a binding credit constraint that prevents parents from borrowing against future child income.” Empirical work supports this mechanism. \citet{kremer2002income} and \citet{de2003inequality} document that countries facing tighter liquidity constraints tend to display higher fertility and lower educational attainment, consistent with liquidity-constrained parents favoring quantity over quality. Cross-country evidence likewise indicates that where financial frictions are more severe, the negative correlation between fertility and schooling is stronger. Theoretically, introducing a borrowing limit can generate multiple steady states: one with low fertility and high investment when incomes suffice to cover quality costs, and another with high fertility and minimal investment when they do not. Policy instruments such as education subsidies or conditional cash transfers effectively relax these constraints, which nudge households toward the low-fertility, high-investment equilibrium. In short, incorporating credit market imperfections deepens the explanatory power of the Q-Q model: economic growth alone may not reduce fertility if households remain too cash-poor to afford schooling, whereas targeted quality-enhancing transfers can catalyze both demographic and human-capital transitions.

A further refinement of the quantity-quality (Q–Q) model considers the question of *who* within the household makes fertility and child investment decisions. The original Beckerian framework adopts a unitary model of the family, and assumes a single utility function and complete agreement between spouses over optimal fertility \( n \) and child quality \( q \). In practice, however, empirical evidence reveals significant heterogeneity in preferences between household members along gender lines. According to \citet{oppenheim1987impact, thomas1990intra}, for instance, men often desire more children than women and may differ in their willingness to invest in each child’s education or health. These discrepancies have motivated game-theoretic models of intra-household bargaining, in which fertility and investment outcomes reflect the relative influence of each parent’s preferences.

@@

The original Q–Q models often used a single catch-all variable for child quality, typically thought of as education or “expenditure per child.” Recent work emphasizes that child quality is multi-faceted, and that parents make trade-offs along several dimensions of investment—cognitive development, health, nutrition, etc. This is salient in developing countries, where basic health and nutrition are pressing concerns alongside schooling.

@@

Beyond survival, early-life health indicators such as birth weight, perceived size at birth, and vaccination status are important dimensions of child quality, especially in low-income settings. When resources are limited, parents may struggle to meet the basic needs of each child. Empirical evidence aligns with the quantity–quality framework: children from larger families often have worse health outcomes at birth and are less likely to receive core immunizations. For example, recent multi-country data show that the share of completely unimmunized children—so-called “zero-dose” children—increases from 10.5% among those with no siblings to 17.2% among those with four or more \citep{costa2024child}. Studies from Southeast Asia also report that higher birth order correlates with lower average birth weight and smaller perceived size at birth. These patterns suggest that as fertility increases, the health of each child may suffer due to more limited parental investment. Studies from Southeast Asia \citep{chen2021population} also report that higher birth order correlates with lower average birth weight and smaller perceived size at birth

Similar patterns emerge in sub-Saharan Africa \citep{bishwakarma2019first} and South Asia, where first-born children experience slower growth once younger siblings arrive—again pointing to intrahousehold trade-offs. Theoretically, one can extend the Becker model with a nutrition production function, where a child’s health \( H \) depends on food \( F \) and medical care \( M \), such that:

@@

Early empirical work in the Philippines provided suggestive support for this theory. For example, an influential study by \citet{horton1986child} used Philippine household data to jointly examine fertility and child nutrition and treated nutritional status as a measure of child quality. Horton found that better-educated mothers and fathers tended to choose smaller families and achieved better-nourished children–evidence of substitution away from “quantity” toward child “quality.” Notably, she observed strong birth-order effects on nutrition (first-born children faring better than later-born), which hint that parents may not evenly distribute quality investments among all children. This early work indicated a quantity-quality (Q–Q) trade-off in Philippine families, though it largely documented correlations rather than definitive causation.

@@

The IV estimates confirmed a significant negative causal effect of higher fertility on educational outcomes. \citet{orbeta2010number} found that each additional child in the household reduced the proportion of school-age children (6–24 years) attending school by roughly 19% of the baseline attendance rate. The trade-off was especially pronounced at higher education levels: for example, the estimated drop in school attendance was about 26% at the secondary level and 57% at the tertiary level for each additional sibling. These are sizable effects, implying that children from large families are substantially less likely to remain in school, presumably due to tighter household budget constraints or diluted parental attention. Moreover, the burden of the trade-off appeared regressive: Orbeta’s results showed much larger schooling deficits from an extra child in poorer households than in richer ones. For instance, in the poorest quintile, an additional sibling reduced school attendance by an estimated 24% (for ages 6–24), compared to a 16% reduction in the richest quintile. This regressive pattern aligns with Becker’s theory that resource constraints bind more tightly for low-income families, which makes the Q–Q trade-off more acute. In summary, Orbeta’s study – the first in the Philippines to account for fertility endogeneity – provides robust evidence that increases in family size cause significant declines in child educational attainment and validates the quantity-quality trade-off in this context. Large family size thus emerges as one mechanism contributing to poverty, by impeding children’s human capital accumulation in the Philippines.

@@

The Philippine experience is echoed in other ASEAN countries, where researchers have probed the quantity-quality trade-off with diverse outcomes and methods. In Vietnam, for example, rapid fertility decline alongside rising education led to questions about a Q–Q mechanism. \citet{anh1998family} documented a negative correlation between family size and children’s school enrollment in Vietnam, though their analysis could not fully establish causality.

@@

Nonetheless, the prevailing evidence from the Philippines and its regional neighbors strongly supports Becker’s quantity-quality conjecture. As families have fewer children, they appear to invest more in each child’s education—investments crucial for human capital development and economic growth. Conversely, high-fertility households risk under-investing per child, which reinforces cycles of poverty and inequality. This literature thus provides an important empirical foundation for policies in the Philippines and ASEAN—from family planning programs to education subsidies—that aim to ease the quantity-quality trade-off and help families achieve both manageable size and better outcomes for the next generation.

@@

This study builds upon the foundational framework of the quantity-quality (Q–Q) trade-off, first articulated in \citet{becker1960economic} and later refined in \citet{becker1973interaction}. In this model, fertility and child quality are jointly determined under a full-income constraint, such that increases in the number of children—without a corresponding expansion in household resources—necessitate a reduction in the average investment per child. While empirical applications of this framework have traditionally focused on educational outcomes, child quality can also be expressed in terms of early-life health, including indicators that emerge at birth or shortly after, such as birth weight, perceived size at birth, and uptake of essential vaccinations.

This analysis focuses specifically on early-life health as a proxy for child quality. These outcomes, unlike cognitive or schooling-related investments, are biologically constrained and occur within a narrow developmental window, during which parental inputs are least substitutable and most consequential. The trade-off is expected to be especially salient in settings where fertility increases exogenously while family resources remain fixed. In such contexts, a sudden increase in sibship size may compel households to spread food, healthcare, and caregiving attention across more dependents, which would potentially lead to diminished health investments and outcomes for each child.

@@

In the presence of a fertility shock, holding \( I \) constant, the average \(\kappa_j\) must fall unless offset by changes in parental behavior or external transfers. If these inputs decline and are critical to early-life development, observable deficits may arise in birth weight or perceived neonatal size, and children may be less likely to receive complete immunization schedules.

The first hypothesis derived from this framework is that an exogenous increase in fertility leads to a decline in per-child health investment, particularly in low-income urban households with little financial slack. A second, related hypothesis posits that this decline results in measurable early-life health disadvantages—specifically, lower birth weight, smaller perceived size at birth, and reduced uptake of key childhood vaccinations. These hypotheses describe a causal pathway through which fertility shocks may contribute to early deprivation and potentially reinforce intergenerational cycles of poor health and limited human capital.

@@

\subsection{Data}

The primary data source for this study is the Philippine Demographic and Health Surveys (DHS), specifically the 1998, 2003, and 2008 rounds. These nationally representative surveys provide detailed child- and household-level information, including maternal characteristics, fertility behavior, and early-childhood health outcomes. Importantly, they are the only Philippine DHS waves with available GPS cluster data, which I use to approximate geographic location and assign respondents to treatment and control groups. Because city-level identifiers are not released in the public-use files, I rely on coarse geographic assignment based on cluster coordinates. Clusters located within or near the administrative boundaries of the City of Manila are classified as “treated,” while clusters located in other parts of the National Capital Region (NCR) serve as the control group. In robustness checks, I restrict the control group to Quezon City or expand it to the rest of NCR excluding both Manila and Quezon City.

The analytic sample includes children under five years of age living with their mothers at the time of the survey. I focus on three early-life health outcomes that reflect both biological endowments and parental investment: (1) birth weight, reported in grams when available; (2) the mother’s subjective assessment of birth size; and (3) vaccination uptake, defined as receipt of key early-childhood immunizations, including BCG, DPT, polio, and measles. These outcomes are consistently available across all three waves and serve as proxies for child quality in the context of the quantity–quality trade-off.

The 1998 DHS is used as the primary pre-policy baseline wave because it captures children born before the implementation of Manila’s contraceptive ban in 2000. The 2008 DHS serves as the post-policy round to capture outcomes for children born well after the ban took effect. I also include the 2003 DHS in the main specification, recognizing that children in this round may have been born shortly before or after the policy’s introduction. I treat 2003 as a transition-period dataset and conduct sensitivity checks to assess the robustness of findings to its inclusion. I exclude the 1993 DHS from the analysis due to the absence of GPS data, which makes it impossible to reliably assign treatment status.

To support the identification strategy, I use DHS summary indicators on contraceptive use disaggregated by region and city. These data confirm a sharp and sustained decline in modern contraceptive use in Manila following the implementation of Executive Order No. 003, relative to surrounding cities, validating the policy as a plausibly exogenous fertility shock. I also considered using data from the Philippine National Nutrition Survey (NNS), which includes rich anthropometric and dietary data. However, the NNS microdata are only available from 2003 onward, which limits their value for pre-policy analysis. As such, the DHS remains the most suitable dataset for this study.

\subsection{Identification Strategy}

To estimate the causal effect of increased fertility on early-childhood health outcomes, I exploit a policy shock generated by Executive Order No. 003, issued in 2000 by the mayor of Manila. This order restricted access to modern contraceptives in all city-run health facilities and remained in effect for nearly a decade. The policy was implemented only in the City of Manila and did not affect neighboring municipalities within the National Capital Region (NCR), which continued to provide modern family planning services. This spatially localized and time-bound restriction created a natural experiment that allows for a quasi-experimental research design. I define the treatment group as children residing in GPS clusters located within or near the Manila boundary, and the control group as children in other NCR municipalities. Given the limitations of the DHS GPS data, which are displaced for confidentiality and do not include exact city identifiers, I use a coarser geographic classification to approximate exposure. Robustness checks vary the control group to include Quezon City only, or the rest of NCR excluding both Manila and Quezon City.

The core of my empirical strategy is a difference-in-differences (DiD) design that compares changes in child outcomes in Manila before and after the ban relative to changes in similar urban areas that were unaffected. The key assumption is that, in the absence of the policy, trends in early-childhood health outcomes would have followed similar trajectories across treatment and control areas. This approach accounts for both fixed differences between cities and nationwide time trends.

To address potential violations of the parallel trends assumption—such as Manila-specific shocks that may have affected all children regardless of family size—I implement a triple-differences strategy that leverages variation in exposure by maternal age. Because fertility declines sharply with age, older women were biologically less likely to respond to the contraceptive ban. For example, mothers aged 45–49 at the time of the policy were near the end of their reproductive years and thus unlikely to experience a fertility increase. By comparing changes in child outcomes across maternal age cohorts within Manila and control cities, I can account for time-varying, location-specific confounders that affect all families, while isolating the policy's effect on family size and its consequences.

I further restrict my main analysis to children born before the policy was implemented. Although these children were not conceived due to the ban, they may still have been indirectly affected by the fertility increase through resource dilution within households. This strategy avoids selection bias related to unobservable differences in the endowments of “marginal children” born because of the policy. As argued by Rosenzweig and Zhang (2009), when parents reinforce rather than compensate for birth endowments, estimates based on older siblings provide a lower bound of the true effect of fertility on child quality. Empirical evidence from the Philippines supports this pattern of reinforcement, as shown by Baguet and Dumas (2015). Taken together, these strategies allow me to estimate the causal impact of fertility shocks on early-life health while addressing key concerns about internal validity and sample composition.

\subsection{Sample Selection}

The analytic sample is drawn from the Philippine Demographic and Health Surveys (DHS), specifically the Kid Recode (KR) files from 1998, 2003, and 2008. I restrict the sample to children under age five (0–59 months) with non-missing information on key outcomes: birth weight (in grams, when available), perceived size at birth, and vaccination status. Because the Manila contraceptive ban was enacted in 2000 and likely implemented unevenly at first, I define children born in 2001 and later as post-policy and those born before 1997 as pre-policy. Children born between 1997–2000 are excluded in the main specification but retained in robustness checks.
